---
title: "COVID Analysis Report"
author: 'Karola Takács, Student_id: 2003898'
date: "29/11/2020"
output:
  html_document: default
  pdf_document: default
editor_options:
  chunk_output_type: console
---

## Introduction ##
Outcome variable: **Number of registered deaths**
Explanatory variable: **Number of registered case**

**Research question**: Which countries are underperforming taken their number of registered cases in terms of number of deaths?

**Variables**

- Population data is coming from The World Bank, which gathers cross-sectional internationally comparable statistical data, measured per country in each year.

- The COVID-related variables originate from the JHU CSSE COVID-19 Dataset, where daily case reports are collected. Variables included in the analysis after the cleaning can be found in the variables.xlsx file.

- The population is all the countries where there are reported COVID-cases. This covers most of the world by now.
Potential issues: not that reliable data reporting in 3rd countries; delay in reporting times and date, so cases might get added to a different date when they actually were discovered, but this has a bigger relvance when the case numbers were still low. Deaths and confirmed cases are not 100% discovered and thus reported in general.

- Selecting observations:  grouped my observations by country_region, then removed all the aggregate countries/regions since those were practically duplicates. Removed those observations which were lacking in population data and also dropped countries where either the y or the x variable was NA. Pppulation could be scaled by dividing it by 1000 (the smallest value for that was San Marino with 33860 inhabitants), but this variable was only used for weghting. There are extreme values for China and India in terms of population, but those are not from error, so I kept them.Lastly I filteres the rows where number of deaths was zero, since taking the ln of zero is not defined.


```{r setup, include=FALSE}

rm(list=ls())
library(knitr)
library(xtable)
library(markdown)
library(tidyverse)
require(scales)
library(lspline)
library(estimatr)
library(texreg)
library(ggthemes)
knitr::opts_chunk$set(echo = FALSE,warning = FALSE, message = FALSE)
df <- read_csv( "https://raw.githubusercontent.com/hushva/DA2_Assignment/master/data/clean/covid_pop_02_11_2020_clean.csv")
rstudioapi::viewer
```

```{r filter}
df <- filter(df,death != 0,)
```


## Summary statistics ##


```{r sum stats}

conf_summary <- df %>%  summarise(
  mean = mean( confirmed ),
  median = median( confirmed ),
  sd = sd( confirmed  ),
  min = min( confirmed ),
  max = max( confirmed ))

death_summary <- df %>%  summarise(
  mean = mean( death ),
  median = median( death ),
  sd = sd( death ),
  min = min( death ),
  max = max( death ))
```

```{r stats table, results='asis'}

```


The distribution of both variables is skewed with a long right tail.

### Transformations ###

* Level-level: quasi linear (little curvy) but the problem is that the observations are grouped to the lower left corner. This skewness was already visible on the distribution histogram above. This is not giving us any useful patterns of association.
* Taking the natural logarithm of either X or Y variables indicates a non-linear pattern. In case of ln(confirmed cases) the distribution is less skewed though, but the shape is closer to exponential or maybe parabola (only a segment of the U shape)
* Log-log transformation: we could say that the the distribution of X and Y much more symmetric and not that many extreme values. It looks almost as a 45°line, but probably more difficult to interpret.
Since I removed observations with zero values for the variable death, log can be applied to it as well.


### Estimating different models ###

- **Linear model**: For countries with 1% more confirmed cases, we would expect the number of deaths to change by 1.02%, on average.
- **Quadratic model**: We can see that the parabola in convex (positive Beta2)
- **PLS**:
  - When comparing observations with number of confirmed cases lass than 5000, number of deaths is 0.81% higher, on     average, for observations with one percent higher number of confirmed cases.
  - When comparing observations with number of confirmed cases over 5000, number of deaths is 1.08% higher, on     average, for observations with one percent higher number of confirmed cases.
- **Weighted linear**: For countries with 10% more confirmed cases we would expect number of deaths to be higher by 9.5% on average.

**R-squared**: initially quite high, the best in case of the weighted model: this model explains 93% of the variation in the outcome variable and only 7% is left for residual variation.


```{r, echo = F, include = FALSE}
# Take both logs
df <- df %>% mutate( ln_death = log( death ),
                     ln_confirmed = log( confirmed ) )

# Make some models:

#     reg1: linear:       ln_death = alpha + beta * ln_confirmed
#     reg2: exponential:  ln_death = alpha + beta_1 * ln_confirmed + beta_2 * ln_confirmed^2
#     reg3: cubic:        ln_death = alpha + beta_1 * ln_confirmed + beta_2 * ln_confirmed^2 + beta_3 * ln_confirmed^3
#     reg4: weighted-ols: ln_death = alpha + beta * ln_gdppc, weights: population

df <- df %>% mutate( ln_confirmed_sq = ln_confirmed^2,
                     ln_confirmed_cb = ln_confirmed^3)
```


Model formula: ln(death) = -3.01 + 0.95*ln(confirmed)
Interpretation:
- Alpha: not interpreted
- Beta: for countries with 10% more confirmed cases we would expect number of deaths to be higher by 9.5% on average.

### Hypothesis testing ###

Our null hypothesis says that beta = 0
Chosen significance level: 0.05
As the p-value is much less than 0.05, we can reject the null hypothesis that beta = 0. Hence there is a significant relationship between the variables.


```{r}
reg5 <- lm_robust(ln_death ~ ln_confirmed, data = df , weights = population)
```


### Residual analysis ###

```{r residuals}
# Get the predicted y values from the model
df$reg5_y_pred <- reg5$fitted.values
# Calculate the errors of the model
df$reg5_res <- df$ln_death - df$reg5_y_pred 

# Find countries with largest negative errors
top5 <- df %>% top_n( -5 , reg5_res ) %>% 
  select( country , ln_death , reg5_y_pred , reg5_res )
# Find countries with largest positive errors
bottom5 <- df %>% top_n( 5 , reg5_res ) %>% 
  select( country , ln_death , reg5_y_pred , reg5_res )
```

```{r res table}
print(top5)
print(bottom5)
```

Countries with the largest negative errors are: Burundi, Iceland, Quatar, Singapore, Sri Lanka. These countries have a lower number of deaths than it was predicted by the regression model.Similarily countries with the largest positive errors are: Bolivia, Ecuador, Iran, Mexico, Yemen. These have higher actual number of deaths than it was predicted by the model. Since the residuals are not that big (though they are ln) probably outliers have not been overlooked. 

## Summary ##



# Appendix #

### Distributions ###
```{r, echo = F,warning=FALSE, message=FALSE, results = 'hide'}
df %>%
  ggplot(aes(x=confirmed)) +
  geom_histogram()+
  theme_gray()+
  labs(x= "Nr. of confirmed cases", y= "Frequency", title = "Figure 1.1 Distribution of confirmed cases")

df %>%
  ggplot(aes(x=death)) +
  geom_histogram()+
  theme_gray()+
  labs(x= "Nr. of deaths", y= "Frequency", title = "Figure 1.2 Distribution of registered number of deaths")
```


### Investigating transformations ### 
```{r, echo = F, warning=FALSE, message=FALSE, results = 'hide'}
# death - confirmed level-level model without scaling
ggplot( df , aes(x = confirmed, y = death)) +
  geom_point() +
  geom_smooth(method="loess")+
  labs(x = "Number of confirmed cases",y = "Number of deaths", title = "Figure 2.1 Level-level transformation")

# death - confirmed level - log model
ggplot( df , aes(x = confirmed, y = death)) +
geom_point() +
  geom_smooth(method="loess")+
  labs(x = "Number of confirmed cases, ln scale",y = "Number of deaths", title = "Figure 2.2 Level-log transformation") +
  scale_x_continuous( trans = log_trans(),  breaks = c(10,50,100,200,500,1000,5000,10000))

# death - confirmed log - level model
ggplot( df , aes(x = confirmed, y = death)) +
  geom_point() +
  geom_smooth(method="loess")+
  labs(x = "Number of confirmed cases",y = "Number of deaths, ln scale", title = "Figure 2.3 Log-level transformation") +
  scale_y_continuous( trans = log_trans() )


# death - confirmed log - log model
ggplot( df , aes(x = confirmed, y = death)) +
  geom_point() +
  geom_smooth(method="loess")+
  labs(x = "Number of confirmed cases, ln scale",y = "Number of deaths, ln scale", title = "Figure 2.4 Log-log transformation") +
  scale_x_continuous( trans = log_trans(), breaks = c(10,50,100,200,500,1000,5000,10000)) +
  scale_y_continuous( trans = log_trans() )
```

## Model estimation ##

```{r reg1, echo = F, warning=FALSE, message=FALSE, results = 'hide'}
# reg1:
reg1 <- lm_robust( ln_death ~ ln_confirmed , data = df , se_type = "HC2" )
# Summary statistics
summary( reg1 )

# Visual inspection:
ggplot( data = df, aes( x = ln_confirmed, y = ln_death) ) + 
  geom_point( color='seagreen', size = 2) +
  geom_smooth( method = lm , color = 'tomato2') +
  labs(x = "Number of confirmed cases, ln scale",y = "Number of deaths, ln scale", title = "Regression 1 - Linear model", caption = "Figure 3.1" )
```

```{r reg2, echo = F, warning=FALSE, message=FALSE, results = 'hide'}
# reg2:
reg2 <- lm_robust( ln_death ~ ln_confirmed + ln_confirmed_sq , data = df )
summary( reg2 )

ggplot( data = df, aes( x = ln_confirmed, y = ln_death ) ) + 
  geom_point( color='seagreen', size = 2) +
  geom_smooth( formula = y ~ poly(x,2) , method = lm , color = 'tomato2' ) +
  labs(x = "Number of confirmed cases, ln scale",y = "Number of deaths, ln scale", title = "Regression 2 - Quadratic  model", caption = "Figure 3.2" )
```

```{r reg3, echo = F, warning=FALSE, message=FALSE, results = 'hide'}
# reg3:
reg3 <- lm_robust( ln_death ~ ln_confirmed + ln_confirmed_sq + ln_confirmed_cb , data = df )
summary( reg3 )

ggplot( data = df, aes( x = ln_confirmed, y = ln_death) ) + 
  geom_point( color='seagreen') +
  geom_smooth( formula = y ~ poly(x,3) , method = lm , color = 'tomato2' ) +
  labs(x = "Number of confirmed cases, ln scale",y = "Number of deaths, ln scale", title = "Regression 3 - Cubic  model", caption = "Figure 3.3")
```

```{r reg4, echo = F, warning=FALSE, message=FALSE, results = 'hide'}
# reg4:
# Regression with piecewise linear spline:
# define the cutoff for number of confirmed cases
cutoff <- 5000
# use a log transformation
cutoff_ln<- log( cutoff )
# Use simple regression with the lspline function

reg4 <- lm_robust(ln_death ~ lspline( ln_confirmed , cutoff_ln ), data = df )
summary( reg4 )
ggplot( data = df, aes( x = ln_confirmed, y = ln_death) ) + 
  geom_point( color='seagreen') +
  geom_smooth( formula = y ~ lspline(x,cutoff_ln) , method = lm , color = 'tomato2' ) +
  labs(x = "Number of confirmed cases, ln scale",y = "Number of deaths, ln scale", title = "Regression 4 - PLS  model", caption = "Figure 3.4" )
```


```{r reg5, echo = F, warning=FALSE, message=FALSE, results = 'hide'}
summary( reg5 )

ggplot(data = df, aes(x = ln_confirmed, y = ln_death)) +
  geom_point(data = df, aes(size=population),  color = 'seagreen4', shape = 18, alpha = 0.6,  show.legend=F) +
  geom_smooth(aes(weight = population), method = "lm", color='tomato2', se = F, size = 0.7) +
  labs(x = "ln(Number of confirmed cases)",y = "ln(Number of death)", title = "Regression 5 - Weighted linear  model", caption = "Figure 3.5" ) +
 # theme_gray() +
  annotate("text", x = 16.047702, y = 12.3532548, label = "USA", size = 4) +
  annotate("text", x = 11.423570, y = 8.4635814, label = "China", size = 4)+
  annotate("text", x = 15.927858,  y = 11.7207279, label = "India", size = 4)
```

```{r echo=FALSE, results='asis'}
tmp <- URLencode(paste(readLines("~/Documents/CEU/Courses/2020_Fall/Mandatory/DA2/DA2_Assignment/out/model_comparison.html"), collapse="\n"))

cat('<iframe src="data:text/html;charset=utf-8,', tmp ,
    '" style="border: none; seamless:seamless; width: 800px; height: 200px"></iframe>')
```


Based on model comparison my chosen model is reg5 - weighted linear

   Substantive: - log-log interpretation works properly with these type of variables
                - magnitude of coefficients are meaningful
   Statistical: - linear is a good approximation
                - very high R-squared and captures variation well
                
```{r hypothesis}
summary( reg5 )

```